Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-11-28 response_to_reviewers.Rmd

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Major comments from Reviewer 1

> Overall, I thought the authors were successful in introducing the big ideas in sports
analytics at a general and non-technical level, and many interesting aspects of the literature
were discussed. However, I think the second part of the title ”Statistical Tools for Their
Investigation” was not properly explored. Several packages are mentioned throughout, and
a brief sketch of a possible analysis workflow is given in Section 5, but I think it lacks sufficient
detail. Beyond that, I think there are some extraneous figures and possible improvements
with the overall flow.

Our revision includes a significant bolstering of Sections 4, 5, 6, and 7. In particular, we have added a section on the use of non-performance data, added a subsection on Elo ratings, added a subsection on baseball tools, expanded and linked the discussion of graphical tools, including information about sports analytics conferences, replaced some weak figures with more relevant substitutes, and added supporting references. We hope these additions will satisfy the reviewer.

@@

> I think this section could use a table of some sort, perhaps something that summarizes the CRAN task view at a high level. Another potential would be a figure which shows a typical workflow of sports analytics as explained in the text.

We have added a table summarizing the sport-specific packages that are available on CRAN.

> Figure 6 doesn't add much to the text; I don't think the reader better understands sports analytics by visualizing many team colors. Instead, it would be better to show  how the theming can lead to a more impactful graphic for sports, i.e. show how that package might be used in a real-world setting.

We thank the reviewer for this excellent suggestion. We have replaced Figure 6 with a new figure that illustrates how the `retrosheet` and `teamcolors` package can be combined to produce a meaningful data graphic. In particular, we note how the use of team colors clarifies what would otherwise be a messy jumble of indistinguishable lines.

@@

> 12, 23: This is not much of a case study.  Perhaps some additional information on the data available, ways it could be used to calculate game state expected valued, win probability, player ratings….

We have connected the case study in chess to the new subsection on Elo ratings and the subsequent subsection on graphical tools. We hope that these explicit connections to additional content will make this subsection feel more substantial. 
In addition, we have added new subsection to this Section with a different case study in baseball about the evolution of data, tools, and statistical methodologies. We hope that this additional content will help to flesh out the value of these tools.

> 12, 37: Theming *teamcolors* is a great *R* package, not sure it is part of the "big ideas" in sports analytics.  A single line in the first paragraph of this section would seem to be better.

We have added examples from the `SportyR` package, which draws playing surfaces, to help flesh out this subsection. We also replaced the previous Figure 6 with a new Figure 6 that illustrates how the use of team colors is particularly impactful for sports-related data graphics.

### 2022-11-28 sports_analytics.Rmd
finishing touches

Doing the work of sports analytics requires computing with data. 
While the sources of sports data are too numerous to list, in Section \ref{sec:tools} we highlight a few computational tools that make this kind of work possible. 
Section \ref{sec:opportunities} lists several opportunities for disseminating work publicly. 
We conclude in Section \ref{sec:conclusion} with a short discussion of some ideas that are not explored in this paper. 
Notably, we omit a treatment of individual player ratings for team sports, since this concept has been covered ably in these pages by @Albert2015, and its inclusion would double the length of this manuscript.
We do, however, discuss individual player ratings in the context of one-person teams (e.g., chess, tennis) in Section \ref{sec:strength}.

@@

wp_plot <- ggplot(x, aes(game_seconds_remaining, wp, color = team)) +
  scale_color_manual(
    values = setNames(miracle$primary, c("home_wp", "away_wp")), 
    labels = miracle$name
  ) +
  geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") +
  geom_vline(
    xintercept = seq(0, 3600, 900), 
    linetype = "dashed", 
    color = "gray"
  ) + 
  geom_line(size = 1.5, alpha = 0.8) +
  labs(
    x = "Time Remaining (seconds)",
    y = "Win Probability",
    color = NULL
  ) +
  theme(legend.position = "top")

@@

The first method for estimating team strength to become popular in sports analytics is a measure of expected winning percentage---often called Pythagorean expectation---developed by @james2003new. 
Later, @miller2007derivation derived the formula as a consequence of assuming that runs (in baseball) are generated by two independent Weibull processes.

@@

Subsequent analysts have tried to find the optimal value of $\alpha$ for various time periods. 
This can be done with a few lines of code, after observing that 
$$
  \frac{X^\alpha}{X^\alpha + Y^\alpha} = \frac{1}{1 + (Y/X)^2}
$$
and fitting a non-linear model (see similar discussion in @baumer2021mdsr). 
Figure \ref{fig:pythag} illustrates the quality of the fit in Major League Baseball since 1954.

@@

```{r pythag, fig.cap="Winning percentages vs. runs scored and runs allowed in baseball, 1954--2021. The navy line represents the expected winning percentage model posited by James, with the exponent value of 2. The gold line shows the same model with an optimal exponent of 1.84."}
data_space +
  stat_function(fun = exp_wpct, args = list(2), size = 2, color = "navy") +
  stat_function(fun = exp_wpct, args = list(optimal), size = 2, color = "gold")
```

@@

Perhaps the most widely-used probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. 
For a pair of players (or teams) $i$ and $j$, let $\Pi_{ij}$ denote the probability that $i$ is preferred to $j$. 
Then the BTM is a logistic regression model with parameters $\beta_i, \beta_j$ such that
$$
\log\left(\frac{\Pi_{ij}}{\Pi_{ji}}\right) = \beta_i - \beta_j \,.
$$
Here, $\exp{\beta_i}$ is often viewed as a representation of team $i$'s ability.

@@

# Non-performance data

@@

## The use of sports betting market data {#sec:betting-market-data}

@@

```{r chess, eval=whoami::username() == 'bsbaumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package. At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}
library(chess)
# Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
game() |> plot()

@@

In drawing together these three big ideas in sports analytics, we have also drawn attention to some computational tools for doing the work and opportunities to showcase that work. 
Our discussion in Section \ref{sec:baseball} shows how the increased resolution of available data has catalyzed new research directions in baseball, but this same dynamic is playing out in all sports. 
It is through these exchanges of ideas, tools, models, and data that analytics moves our collective understanding of sports forward.

>>>>

Doing the work of sports analytics requires computing with data. 
While the sources of sports data are too numerous to list, we discuss the use of non-performance data (most notably, sports market betting data) in Section \ref{sec:betting}, and in Section \ref{sec:tools} we highlight a few computational tools (including a table of R packages) that make this kind of work possible. 
Section \ref{sec:opportunities} lists several opportunities for disseminating work publicly. 
We conclude in Section \ref{sec:conclusion} with a short discussion of some ideas that are not explored in this paper. 
Notably, we omit a treatment of individual player ratings for team sports, since this concept has been covered ably in these pages by @Albert2015, and its inclusion would double the length of this manuscript.
We do, however, discuss individual player ratings in the context of one-person teams (e.g., chess, tennis) in Section \ref{sec:strength}.

@@

wp_plot <- ggplot(x, aes(game_seconds_remaining, wp, color = team)) +
  scale_color_manual(
    values = setNames(miracle$primary, c("home_wp", "away_wp")), 
    labels = miracle$name
  ) +
  geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") +
  geom_vline(
    xintercept = seq(0, 3600, 900), 
    linetype = "dashed", 
    color = "gray"
  ) + 
  geom_line(linewidth = 1.5, alpha = 0.8) +
  labs(
    x = "Time Remaining (seconds)",
    y = "Win Probability",
    color = NULL
  ) +
  theme(legend.position = "top")

@@

A simple method for estimating team strength that has become popular in sports analytics is expected winning percentage---often called Pythagorean expectation---developed by @james2003new. 
Later, @miller2007derivation derived the formula as a consequence of assuming that runs (in baseball) are generated by two independent Weibull processes.

@@

Subsequent analysts have tried to find the optimal value of $\alpha$ for various time periods. 
This can be done with a few lines of code, after observing that 
$$
  \frac{X^\alpha}{X^\alpha + Y^\alpha} = \frac{1}{1 + (Y/X)^2}
$$
and fitting a non-linear model (see similar discussion in @baumer2021mdsr). 
Figure \ref{fig:pythag} illustrates the quality of the fit in Major League Baseball since 1954, where the optimal value of $\alpha$ is 1.84.

@@

```{r pythag, fig.cap="Winning percentages vs. runs scored and runs allowed in baseball, 1954--2021. The navy line represents the expected winning percentage model posited by James, with the exponent value of 2. The gold line shows the same model with an optimal exponent of 1.84."}
data_space +
  stat_function(fun = exp_wpct, args = list(2), linewidth = 2, color = "navy") +
  stat_function(fun = exp_wpct, args = list(optimal), linewidth = 2, color = "gold")
```

@@

Perhaps the most widely-used probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. 
For a pair of players (or teams) $i$ and $j$, let $\Pi_{ij}$ denote the probability that $i$ is preferred to $j$. 
Then the BTM is a logistic regression model with parameters $\beta_i, \beta_j$ such that
$$
\log\left(\frac{\Pi_{ij}}{\Pi_{ji}}\right) = \beta_i - \beta_j \,.
$$
Here, $\exp{(\beta_i)}$ is often viewed as a representation of team $i$'s ability.

@@

# Non-performance data {#sec:betting}

@@

## The use of betting market data for sports analytics {#sec:betting-market-data}

@@

```{r chess, eval=whoami::username() == 'bbaumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package. At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}
library(chess)
# Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
game() |> plot()

@@

In drawing together these three big ideas in sports analytics, we have also drawn attention to new uses of sports betting market data, some computational tools for doing sports analytics work, and opportunities to showcase that work. 
Our discussion in Section \ref{sec:baseball} shows how the increased resolution of available data has catalyzed new research directions in baseball, but this same dynamic is playing out in all sports. 
It is through these exchanges of ideas, tools, models, and data that analytics moves our collective understanding of sports forward.
